A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-03-03 papers/contact/paper.tex

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  The chain and association contributions to the SAFT
free energy are of particular interest for this paper, since each uses
as input the contact value of the correlation function.
%
Yu and Wu introduced in 2002 a functional for the association term of
the free energy, which included a functional for the contact value of
the correlation function (see
Section~\ref{sec:yuwu})\cite{yu2002fmt-dft-inhomogeneous-associating},
which has subsequently been used in other
functionals\cite{fu2005vapor-liquid-dft, bryk2006density}.
%% Each of these terms has some dependence on the
%% inhomogeneity in the density.  The ideal gas is purely local, and in
%% fact must be the only purely local term in the free energy in order
%% for the functional to satisfy the contact-value theorem.  The
%% hard-sphere contribution contribution to the free energy has been
%% thoroughly studied, and is well approximated by the White Bear version
%% of the Fundamental-Measure Theory (FMT)
%% functional\cite{roth2002whitebear}.
Two functionals for the chain contribution have recently been
introduced\cite{bryk2006density, gross2009density}, one which uses the
contact value of the correlation function of Yu and
Wu\cite{bryk2006density}, while the other introduces a new functional
for the contact value of the correlation
function (see Section~\ref{sec:gross})\cite{gross2009density}.
%% The dispersion term is long-range and thus has significant
%% dependence on the density distribution, but because its relatively
%% weak position dependence, is amenable to mean-field approximations.
Given these different approaches, it seems worthwhile to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.

@@

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
$2R$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
%
\fixme{The following bit is awkward.}
%
In a system with an inhomogeneous density, we seek a \emph{local}
contact value of the correlation function.  There are two reasonable
options for defining such a local function:  a symmetric and an
asymmetric formulation.

The first option treats the two particles that are in contact
symmetrically:
\begin{align}
  g^a_\sigma(\rr) &= \frac{n^a_\textit{contact}(\rr)}{n_a(\rr)}
\end{align}
\begin{align}
  n^a_\textit{contact}(\rr) &= \frac1{4\pi R^2}\int n^{(2)}(\rr - \rr', \rr + \rr')\delta(|\rr'| -
  R)d\rr'
\end{align}
\begin{align}
  n_a(\rr) &= \frac{1}{4\pi R^2}\int n(\rr')\delta(|\rr-\rr'|-R) d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_a$ is the same as the density $n_0$
defined in Fundamental Measure Theory (FMT), and seems ideal for
treating touching spheres.

The second option is to treat one of the two touching spheres
specially:
\begin{align}
  g^b_\sigma(\rr) &= \frac{n^b_\textit{contact}(\rr)}{n_b(\rr)} \\
  n^b_\textit{contact}(\rr) &= \frac1{16\pi R^2}\int n^{(2)}(\rr, \rr + \rr')
  \delta(|\rr'| - 2R)d\rr' \\
  n_b(\rr) &= \frac1{16\pi R^2}\int n(\rr') \delta(|\rr-\rr'|-2R) d\rr'
\end{align}
In this case, we find the average number of contacts for a sphere at a
given position.

In the process of defining these two averaged correlation functions,
we also defined four averaged densities, two \emph{contact
  densities}, $n^a_\textit{contact}(\rr)$ and $n^b_\textit{contact}(\rr)$,
and two \emph{contacting densities}, $n_a(\rr)$ and $n_b(\rr)$.  The
former are the density of spheres that are in contact with each other,
while the latter are the density of spheres available to be in
contact.  In the homogeneous limit, $n_a = n_b = n$ and
$n^a_\textit{contact} = n^b_\textit{contact} = g(\sigma)n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

Yu and Wu introduce an approximation for
$g^a$\cite{yu2002fmt-dft-inhomogeneous-associating}, which will be
discussed in Section~\ref{sec:yuwu}.  Gross introduced an
approximation for $g^b$\cite{gross2009density}, which will be
introduced in Section~\ref{sec:gross}.  We will also introduce our own
approximation to $g^b$ which is based on the assumption of
thermodynamic consistency within FMT, in
Section~\ref{contact-at-sphere}.  We will report on the accuracy of
these approximations by comparing with Monte Carlo simulations of the
inhomogeneous hard-sphere fluid.  We will focus on the contact
densities, rather than the correlation function itself, because these
are simpler, and because they are the quantity that you want to
``average''.  i.e. if you were to average $g$ itself, you'd want to
weight that average based on the density, which would mean averaging
$n^{(2)}$.

@@

  \begin{align}
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    -\frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}\left(
    \frac{1}{(1-n_3)^2} + 2\frac{\ln(1-n_3)}{n_2^2(1-n_3)}
    \right)
    \\
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')} &=
    \frac{n_1}{1-n_3}
    + 3(n_2^2 - \mathbf{n}_{V2}\cdot\mathbf{n}_{V2})\frac{n_3 +
      (1-n_3)^2\ln(1-n_3)}{
      36\pi n_3^2(1-n_3)^2
    }
    \\
    \frac{\delta A_{HS}}{\delta n_1(\mathbf{r}')} &= \frac{n_2}{1-n_3}
    \\
    \frac{\delta A_{HS}}{\delta n_0(\mathbf{r}')} &=
    -\frac{1}{4\pi R^2} \ln\left( 1 - n_3\right) \\
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} &=
    \frac{\mathbf{n}_{V1}}{1-n_3}
    - 6 n_2 \mathbf{n}_{V2} \frac{n_3 +
      (1-n_3)^2\ln(1-n_3)}{
      36\pi n_3^2(1-n_3)^2
    }    \\
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V1}(\mathbf{r}')} &=
    \frac{\mathbf{n}_{V2}}{1-n_3} \\
    \frac{\delta A_{HS}}{\delta R(\mathbf{r}')} &=
    \frac{n_2}{2\pi R^3} \ln\left( 1 - n_3\right)
    - \frac1{R} \frac{n_1n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
  \end{align}

  \begin{align}
    \frac{\delta n_3(\mathbf{r}')}{\delta R(\mathbf{r})} &=
    n(\mathbf{r})\delta(|\mathbf{r}-\mathbf{r}'| - R) \\
    \frac{\delta n_2(\mathbf{r}')}{\delta R(\mathbf{r})} &=
    n(\mathbf{r}) \delta'(|\mathbf{r}-\mathbf{r}'| - R) \\
    \frac{\delta \mathbf{n}_{V2}(\mathbf{r}')}{\delta R(\mathbf{r})} &=
    -\frac{\delta}{\delta R(\mathbf{r})} \mathbf{\nabla}' n_3(\mathbf{r}')
    \\
    &= -\mathbf{\nabla}'
    \left(n(\mathbf{r})\delta(|\mathbf{r}-\mathbf{r}'| - R) \right)
  \end{align}
  These give us some basic tools with which to perform derivatives.

>>>>

Constructing a classical density functional based on SAFT involves
rewriting each term in the free energy as a functional of an
inhomogeneous density.  The SAFT free energy consists of four terms:
\begin{align}
  A_\textit{SAFT} &= A_\textit{ideal} + A_\textit{HS} + A_\textit{chain} + A_\textit{disp} + A_\textit{assoc}
\end{align}
where $A_\textit{ideal}$ is the ideal-gas free energy, $A_\textit{HS}$
is the excess free energy of a hard-sphere fluid of monomers,
$A_\textit{chain}$ is the contribution to the free energy from the
chaining effect in a polymeric fluid, $A_\textit{disp}$ is the
dispersion contribution to the free energy, and $A_\textit{assoc}$ is
the free energy contribution from \emph{association}, which is to say,
hydrogen bonding.  The chain and association contributions to the SAFT
free energy are of particular interest for this paper, since each uses
as input the contact value of the correlation function.
%
Yu and Wu introduced in 2002 a functional for the association term of
the free energy, which included a functional for the contact value of
the correlation function (see
Section~\ref{sec:yuwu})\cite{yu2002fmt-dft-inhomogeneous-associating},
which has subsequently been used in other
functionals\cite{fu2005vapor-liquid-dft, bryk2006density}.
%% Each of these terms has some dependence on the
%% inhomogeneity in the density.  The ideal gas is purely local, and in
%% fact must be the only purely local term in the free energy in order
%% for the functional to satisfy the contact-value theorem.  The
%% hard-sphere contribution contribution to the free energy has been
%% thoroughly studied, and is well approximated by the White Bear version
%% of the Fundamental-Measure Theory (FMT)
%% functional\cite{roth2002whitebear}.
Two functionals for the chain contribution have recently been
introduced\cite{bryk2006density, gross2009density}, one which uses the
contact value of the correlation function of Yu and
Wu\cite{bryk2006density}, while the other introduces a new functional
for the contact value of the correlation
function (see Section~\ref{sec:gross})\cite{gross2009density}.
%% The dispersion term is long-range and thus has significant
%% dependence on the density distribution, but because its relatively
%% weak position dependence, is amenable to mean-field approximations.
Given these different approaches, it seems worthwhile to examine this
property of the hard-sphere fluid through direct simulation, in order
to establish the advantages and disadvantages of each approach.
\fixme{

Jeff's Notes:
A novice (me) is confused by the use of 'contact value of correlation function' throuhout this introduction while throughout the rest of the paper we refer to the 'contact density'.  Pehaps the audince will understand that they're interchangable (if they are) but this was confusing to me.  Also, I would think I'd like to see something like:

We first discuss two possibilities for the definition of contact density.  The first involves a more symmetric interpretation and so we call it the 'S case'.  The second involves a sphere-centered, assymetrical interpretation and so we call it the 'A case'.  We present where in the SAFT free energy formulation these contact densities will contribute.  We next present a thermodynamic method that we will use to calculate the above contact densities.  This method involves the use of surface pressure and the minimization of the excess free energy refered to above.  We use the White Bear and White Bear Mark II formulations of these free energy terms and take neccesary derivitives in the concpetual spirit of the seperately defined S an A case contact densities.  Finally we compare the resulting contact densities with those of Gross, those of Yu and Wu, and as well with our wn monte carlo simulations.}

@@

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
$2R$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
%
\fixme{The following bit is awkward.}
%
In a system with an inhomogeneous density, we seek a \emph{local}
contact value of the correlation function.  There are two reasonable
options for defining such a local function:  a symmetric (which we refer to as S) and an
asymmetric (which we refer to as A) formulation.

For the S case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{n^S_\textit{contact}(\rr)}{n_S(\rr)}
\end{align}
\begin{align}\label{eq:n-contactS}
  n^S_\textit{contact}(\rr) &= \frac1{4\pi R^2}\int n^{(2)}(\rr - \rr', \rr + \rr')\delta(|\rr'| -
  R)d\rr'
\end{align}
\begin{align}
  n_S(\rr) &= \frac{1}{4\pi R^2}\int n(\rr')\delta(|\rr-\rr'|-R) d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_S$ is the same as the density $n_0$
defined in Fundamental Measure Theory (FMT), and seems ideal for
treating touching spheres.

And for the A case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{n^A_\textit{contact}(\rr)}{n_A(\rr)} \\
  n^A_\textit{contact}(\rr) &= \frac1{16\pi R^2}\int n^{(2)}(\rr, \rr + \rr')
  \delta(|\rr'| - 2R)d\rr' \\
  n_A(\rr) &= \frac1{16\pi R^2}\int n(\rr') \delta(|\rr-\rr'|-2R) d\rr'
\end{align}
In this case, we find the average number of contacts for a sphere at a
given position.

In the process of defining these two averaged correlation functions,
we also defined four averaged densities, two \emph{contact
  densities}, $n^S_\textit{contact}(\rr)$ and $n^A_\textit{contact}(\rr)$,
and two \emph{contacting densities}, $n_S(\rr)$ and $n_A(\rr)$.  The
former are the density of spheres that are in contact with each other,
while the latter are the density of spheres available to be in
contact.  In the homogeneous limit, $n_S = n_A = n$ and
$n^S_\textit{contact} = n^A_\textit{contact} = g(\sigma)n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

Yu and Wu introduce an approximation for
$g^S$\cite{yu2002fmt-dft-inhomogeneous-associating}, which will be
discussed in Section~\ref{sec:yuwu}.  Gross introduced an
approximation for $g^A$\cite{gross2009density}, which will be
introduced in Section~\ref{sec:gross}.  We will also introduce our own
approximation to $g^A$ which is based on the assumption of
thermodynamic consistency within FMT, in
Section~\ref{contact-at-sphere}.  We will report on the accuracy of
these approximations by comparing with Monte Carlo simulations of the
inhomogeneous hard-sphere fluid.  We will focus on the contact
densities, rather than the correlation function itself, because these
are simpler, and because they are the quantity that you want to
``average''.  i.e. if you were to average $g$ itself, you'd want to
weight that average based on the density, which would mean averaging
$n^{(2)}$.

@@

  \begin{align}
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    - \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}\left(
    \frac{n_3^2-5n_3+2}{n_3^3(1-n_3)^3} + 2\frac{\ln(1-n_3)}{n_3^3}
    \right)
    \\
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')} &=
    \frac{n_1}{1-n_3}
    + 3(n_2^2 - \mathbf{n}_{V2}\cdot\mathbf{n}_{V2})\frac{n_3 +
      (1-n_3)^2\ln(1-n_3)}{
      36\pi n_3^2(1-n_3)^2
    }
    \\
    \frac{\delta A_{HS}}{\delta n_1(\mathbf{r}')} &= \frac{n_2}{1-n_3}
    \\
    \frac{\delta A_{HS}}{\delta n_0(\mathbf{r}')} &=
    -\ln\left( 1 - n_3\right) \\
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} &=
    \frac{\mathbf{n}_{V1}}{1-n_3}
    - 6 n_2 \mathbf{n}_{V2} \frac{n_3 +
      (1-n_3)^2\ln(1-n_3)}{
      36\pi n_3^2(1-n_3)^2
    }    \\
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V1}(\mathbf{r}')} &=
    \frac{\mathbf{n}_{V2}}{1-n_3} \\
    \frac{\delta A_{HS}}{\delta R(\mathbf{r}')} &=
    \frac{n_2}{2\pi R^3} \ln\left( 1 - n_3\right)
    - \frac1{R} \frac{n_1n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
  \end{align}

*******************************************************************************8
\fixme{The following is the same calculation for WBm2.  Not sure if you want these in the appendix or in the paper at all?}

\begin{align} 
A_{HS} &= \int \left\{
    -n_0 \ln\left(1 - n_3\right)
    + (1 + \frac{1}{9}n_3^2 \phi_2(n_3))   \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (1 - \frac{4}{9}n_3\phi_3(n_3))\frac{(n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})}{24\pi (1-n_3)^2}
    \right\}
\end{align}
 where
 \begin{align}
   \phi_2 &= (6n_3 - 3n_3^2 + 6(1-n_3)\ln(1-n_3))\frac{1}{n_3^3}\\
   \phi_3 &= (6n_3 - 9n_3^2 + 6n_3^3 + 6(1-n_3)^2 \ln(1-n_3))\frac{1}{4n_3^3}
 \end{align}
For the following derivitives we have
\begin{align}
   \frac{\delta \phi_2}{\delta n_3} &= ((-6+4n_3)\ln(1-n_3) - 6n_3 + n_3^2]\frac{3}{n_3^4}\\
   \frac{\delta \phi_3}{\delta n_3} &= (-6n_3 + 5n_3^2 + (-2n_3^2 + 8n_3 -6)\ln(1-n_3))\frac{3}{4n_3^4}
 \end{align}
And using these
\begin{align}
  \frac{\delta A_{HS}}{\delta n_3} &= \psi _1 + \psi _2 + \psi _3
\end{align}
where
\begin{align}
  \psi_1 &= \frac{n_0}{1-n_3}\\
  \psi_2 &= (\frac{n_1n_2 - \mathbf{n}_1\cdot \mathbf{n}_2}{1-n_3})(\frac{2}{9}n_3\phi_2 + \frac{1}{9}n_3^2\frac{\delta \phi_2}{\delta n_3}) + (1 + \frac{1}{9}n_3^2\phi_2)(\frac{n_1n_2 - \mathbf{n}_1\cdot \mathbf{n}_2}{(1-n_3)^2})\\
  \psi_3 &= (\frac{n_2^3 - 3n_2\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1-n_3)^2})(-\frac{4}{9}\phi_3 - \frac{4}{9}n_3\frac{\delta \phi_3}{\delta n_3}) + 2(1 - \frac{4}{9}n_3\phi_3)(\frac{n_2^3 - 3n_2\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1-n_3)^3}) 
\end{align}
We also have
\begin{align}
  \frac{\delta A_{HS}}{\delta n_2} &= (1 + \frac{1}{9}n_3^2\phi_2(n_3))\frac{n_1}{1-n_3} + (1 - \frac{4}{9}n_3\phi_3)(\frac{3n_2^2 - 3\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1 - n_3)^2})\\
  \frac{\delta A_{HS}}{\delta n_1} &= (1 + \frac{1}{9}n_3^2\phi_2)\frac{n_2}{1 - n_3}\\
  \frac{\delta A_{HS}}{\delta n_0} &= -\ln(1 - n_3)\\
  \frac{\delta A_{HS}}{\delta \mathbf{n}_1} &= -(1 + \frac{1}{9}n_3^2 \phi_2)\frac{\mathbf{n}_2}{1 - n_3}\\
  \frac{\delta A_{HS}}{\delta \mathbf{n}_2} &= -(1 + \frac{1}{9}n_3^2\phi_2)\frac{\mathbf{n}_1}{1 - n_3} 
  - (1 - \frac{4}{9}n_3\phi_3)\frac{6\mathbf{n}_2}{24\pi(1 - n_3)^2}
\end{align}

****************************************************************************************\\
Jeff's understanding as of 02/22/2012

  In order to motivate our particular forms for $\frac{\delta n_3^{S} (\mathbf{r}')}{\delta R(\mathbf{r})}$, $\frac{\delta n_2^{S}(\mathbf{r}')}{\delta R(\mathbf{r})}$, etc. we interpret the contact density in the $S$ case as follows: 
  Given that there is a sphere just touching the point $\mathbf{r}'$, $n_{Contact}^{S}(\mathbf{r}')$ is the density
  of spheres that are also touching that point so that they are touching the given sphere.  We can see this in the integral in eq \ref{eq:n-contactS},where we sum over the double densities of spheres that are positioned opposite eachother and touching at the point $\mathbf{r}$ and then divide
  by $n_0$, the average density of a single sphere touching the point.  Thus we have the density of contact at $\mathbf{r}$, where the
  contact between the two spheres actually takes place at $\mathbf{r}$.

Involved in the terms in our free energy functional are derivatives of density functions with respect to $R$.  We take these integrals differently when dealing with the S and the A cases. In both cases we wish to calculate how the density function changes when a 'certain amount of sphere', at some point, is changed.  The A case is conceptually simple.  We say that a changing spherical radius at a point $\mathbf{r}$ represents the outward expansion, by a distance $\delta R$, of the shell of any sphere that is centered at $\mathbf{r}$.  In the case of S we instead expand a multitude of spheres, all of them positioned around and just touching the point $\mathbf{r}$.  When their radii expand each does not expand outward in all directions as in the case of A, rather they each expand outward at only the point $\mathbf{r}$.  One can imagine a shell that touches $\mathbf{r}$ that has an infinitesimal bit of shell area protruding outward at $\mathbf{r}$ while the rest of its shell remains at radius $R$.  When we expand each sphere that touches $\mathbf{r}$ in this peculiar fashion, we have in the end the same amount of shell area protruding outward as we would if we had simply expanded one sphere in fashion of case A.  Also the protrusions, once added up for all spheres sourrounding $\mathbf{r}$, will yield a symmetry in their direction of protrusion.  Each protruding sphere bit will expand outward from its center and all directions will be covered uniformly.

  For the derivation of $n_{3}$ in the symmetric case:

\begin{align}  
  n_3^{S}(\mathbf{r}') &= \int n(\mathbf{r}'') \Theta(\left|\mathbf{r}' - \mathbf{r}''\right| -R(\mathbf{r}')) d\mathbf{r}''\\
  \frac{\delta n_3^{S} (\mathbf{r}')}{\delta R(\mathbf{r})} &=
  \int n (\mathbf{r}'') \delta(|\mathbf{r}' - \mathbf{r}''| - R(\mathbf{r}')) \delta(\mathbf{r}'-\mathbf{r}) d\mathbf{r}''
\end{align}

  $\delta R(\mathbf{r})$ in this equation refers to an expansion, in the fashion described above, of all the spheres that just touch $\mathbf{r}$.  We can see that if the radii change in this fashion, the only point at which $n_3(\mathbf{r}')$ will be affected is at $\mathbf{r}' = \mathbf{r}$, so that $\delta n_3(\mathbf{r}')$ is zero everywhere except where $\mathbf{r}' = \mathbf{r}$.  Thus we have the 2nd $\delta$ function in the lower integral.

This term in the free energy functional, once integrated, takes an interesting form:

\begin{align}
  \int \frac{\delta A_{HS}}{\delta n_3^{S}(\mathbf{r}')} 
  \frac{\delta n_3^{S}(\mathbf{r}')}{\delta R(\mathbf{r})} d\mathbf{r}' 
  &= \int \frac{\delta A_{HS}}{\delta n_3^{S}(\mathbf{r}')} \int n (\mathbf{r}'') \delta(|\mathbf{r}' - \mathbf{r}''| - R(\mathbf{r}')) \delta(\mathbf{r}'-\mathbf{r}) d\mathbf{r}'' d\mathbf{r}' \\
  &= \frac{\delta A_{HS}}{\delta n_3^{S}(\mathbf{r}')} \int n (\mathbf{r}'') \delta(|\mathbf{r} - \mathbf{r}''| - R(\mathbf{r})) d\mathbf{r}''\\
  &= \frac{\delta A_{HS}}{\delta n_3^{S}(\mathbf{r}'')} n_2(\mathbf{r})\\
\end{align}

 $n_2(\mathbf{r}')$ in the case of S in calculated similarly:
\begin{align}
  n_2^{S}(\mathbf{r}') &= \int n(\mathbf{r}'') \delta(|\mathbf{r}' - \mathbf{r}''| + R(\mathbf{r}'))d\mathbf r''\\
    \frac{\delta n_2^{S}(\mathbf{r}')}{\delta R(\mathbf{r})} &= -\int n(\mathbf{r}'') 
  \delta'(|\mathbf{r}'-\mathbf{r}''| - R(\mathbf{r}')) \delta(\mathbf{r}-\mathbf{r}') d\mathbf{r}''
\end{align}

For $\mathbf{n}_{V2}(\mathbf{r}')$ in the case of S we take the gradient of $n_3(\mathbf{r}')$ with respect to $\mathbf{r}'$: 
\begin{align}
  \mathbf{n}_{V2}^{S}(\mathbf{r}') &= \int n(\mathbf{r}'') \delta(|\mathbf{r}' - \mathbf{r}''| - R(\mathbf{r}'))
  \frac{(\mathbf{r}' - \mathbf{r}'')}{|\mathbf{r}' - \mathbf{r}''|} d \mathbf{r}''\\
  \frac{\delta \mathbf{n}_{V2}^{S}(\mathbf{r}')}{\delta R(\mathbf{r})} &= \delta(\mathbf{r} - \mathbf{r}')
  \int n(\mathbf{r}'') \delta'(|\mathbf{r}' - \mathbf{r}''| - R(\mathbf{r}')) 
  \frac{(\mathbf{r} - \mathbf{r}'')}{|\mathbf{r} - \mathbf{r}''|} d\mathbf{r}''
\end{align}

  In the case of A, the changing radius we refer to is the radius of a sphere that is centered at $\mathbf{r}$ and it expands in a more conventional fashion.  The $n_3$ values that will be affected are those which are a distance $R$ fom $\mathbf{r}$.

\begin{align}
  n_3^{A}(\mathbf{r}') &= \int n(\mathbf{r}'') \Theta(\left|\mathbf{r}' - \mathbf{r}''\right| -R(\mathbf{r}'')) d\mathbf{r}''\\ 
  \frac{\delta n_3^{A} (\mathbf{r}')}{\delta R(\mathbf{r})} &=
  \int n (\mathbf{r}'') \delta(|\mathbf{r}' - \mathbf{r}''| - R(\mathbf{r}'')) \delta(\mathbf{r}-\mathbf{r}'') d\mathbf{r}'' \\
   &= n (\mathbf{r}) \delta(|\mathbf{r} - \mathbf{r}'| - R(\mathbf{r}))
\end{align}

And likewise:
\begin{align}
  n_2^{A}(\mathbf{r}') &= \int n(\mathbf{r}'') \delta(|\mathbf{r}' - \mathbf{r}''| - R(\mathbf{r}'')) d \mathbf{r}''\\
  \frac{\delta n_2^{A}(\mathbf{r}')}{\delta R(\mathbf{r})} &= -n(\mathbf{r}) \delta'(|\mathbf{r}' - \mathbf{r}| - R)\\
 \end{align}

 \fixme {
And then there's a lot with taking the derivative of the delta for the case of $n_{2}^{S}(\mathbf{r}')$.  Not sure if we'll use any of this, but wrote down some equations that Roundy wrote out.  So can use the terms at least if going to write about this - I'm just copying it in for now (probably a lotof mistakes)

\begin{align}
  p(\mathbf{r}) &= \frac{\delta A}{\delta n_2(\mathbf{r}')} \frac{\delta n_2(\mathbf{r}')}{\delta R(\mathbf{r})} d\mathbf{r}'\\
  &= \frac{\delta A}{\delta n_2(\mathbf{r})} - \int n(\mathbf{r}'') \delta' (|\mathbf{r} - \mathbf{r}''| - R) d\mathbf{r}''
  \int n (\mathbf{r} +\mathbf{r}'') \delta'(|\mathbf{r}'''| - R) d \mathbf{r}'''\\
  &= \int n(\mathbf{r} + \mathbf{r}''') \delta'(|\mathbf{r}'''| - R) r'''^2 cos\theta d \theta''' dr'''\\
  du = \delta'(\mathbf{r}''' - R) d \mathbf{r}'''     v = n(\mathbf{r} - \mathbf{r}'')r'''\\
   u = \delta(\mathbf{r}''' - R)                       
   dv = \mathbf{\nabla} n(\mathbf{r} + \mathbf{r}'') \cdot \frac{\mathbf{r}'''}{r'''} + 2n(\mathbf{r}' - \mathbf{r}'')r'''\\
  &= -\int(\mathbf{\nabla}n(\mathbf{r}' + \mathbf{r}'') \cdot \mathbf r'''(hat)r'''^2 dr''' 
  + 2n(\mathbf{r} + \mathbf{r}''')r''') dr''' \delta(\mathbf{r} - R) dcos\theta''' d\theta''\\
  &= - \int \delta(\mathbf{r}''' - R) d\mathbf{r}''' 
  (\mathbf{\nabla}n(\mathbf{r} + \mathbf{r}''') \cdot \mathbf{r}''' (hat) + \frac{2n(\mathbf{r} + \mathbf{r}'')}{\mathbf{r}'''})\\
  &= \frac{2n_2(\mathbf{r})}{R} + 
  \int \delta(|\mathbf{r}'''| - R) \mathbf{\nabla}n(\mathbf{r} + \mathbf{r}''') \cdot \mathbf{r}''' (hat) d \mathbf{r}'''\\
\end{align} 
}
********************************************************************************\\

These give us some basic tools with which to perform derivatives.
